The Role of Advection Versus Diffusion in Intraventricular Transport

### 2025-02-24 paper/main.tex
discussiond and conclusion

\subsection{The role of motile cilia in intraventricular flow and transport}
Our work has studied CSF flow and solute distribution
in a relatively small and simple brain model, which is devoid of
active CSF secretion and CSF absorption. Even though there is
a clear conservation of the ventricular system across vertebrates,
there are also important considerations to take. These include (but are not limitied to)
ventricular size and morphology, properties of ciliated cells,
animal posture, and magnitude of velocities. For instance, a large
volume would require an extensive amount of time for solutes to distribute
if based solely on diffusion. Besides, the presence of multiple cilia
instead of a single cilium per cell -- the former can be found in
mammalian ependymal cells -- will generate much stronger velocities.
Hence advection may be more dominant in those situations.
However, in large brains such as the human brain, cilia beating will
only contribute to flow in localized regions of very high proximity to the ventricular walls,
while blood vessel pulsations, respiration and CSF secretion will generate a
pulsatile bulk flow in the central regions of the ventricles~\cite{Siyahhan2014FlowVentricles}.
Based on this, cilia probably play an important role in regulating the
amount of substances close to the ventricular walls, similar to coral
reefs, where it has been observed that cilia are primary regulators of
oxygen amounts at the coral surface~\cite{Pacherres2022CiliaryProduction}.
But for zebrafish embryos, the ventricles are so small that the flow
generated by ciliary motion is of the scale of the geometry, and thus not
only contributes to near-wall flow dynamics, but flow in the entire ventricular system.
To date, motile-cilia deficiency has been associated with altered brain function
in the zebrafish and ventricular enlargement in zebrafish, rodents and humans.
Yet the molecular mechanisms leading to these alterations are poorly understood.
We trust that our model sets the basis for future work addressing these open questions.
We are also confident that our model can be adjusted and implemented to fit
microscopic and macroscopic parameters for larger brains, once these parameters are acquired.

\lyng{HH: Refer to Eichele et al. article discussion about role of cilia.}

\subsection*{Limitations and further work}
The model presented has certain limitations. Future work could consider:
\begin{itemize}
    \item spatial heterogeneity of the cilia, how this affects flow and transport
    \item time-varying cilia?
    \item improved advection-diffusion equation BCs
    \item Modeling deformation of the ventricles
    and investigating how the arising flow compares with pulsatile flow observed in experiments
    \item FSI: CSF flow coupled to 1D/2D/3D cilia structures.
\end{itemize}

\fixme{Based on our conclusion that advection drives transport of larger proteins, could
we make some bold claims about advective transport of exosomes/extracellular vesicles in the
ventricles, and the importance of this?}

\section{Conclusion}
We have presented a numerical model
of cerebrospinal fluid flow and solute transport in
zebrafish brain ventricles based on finite element methods.
A realistic brain ventricles geometry from medical imaging was used to perform a
computational study with the presented model. The flow model reproduces
the experimentally observed flow features of cilia-mediated cerebrospinal fluid
flow in zebrafish brain ventricles. Coupling the flow to the transport model,
we estimated how slowly or rapidly molecules of varying size are transported in
the ventricles. Simulations indicate that cilia-driven advection is important
for transport of larger molecules. 
The numerical model was validated with experiments, measuring the movement of
a photoconverted fluorescent protein in zebrafish brain ventricles,
both for healthy animals and mutant animals with ciliary dysfunction.
We observed good alignment between computational and experimental results.
Experimental results indicate that cilia only affect the local distribution of small proteins,
while most of the transport is governed by diffusion.

@@

Further steps depend on the boundary conditions applied,
which is different for the three model versions we consider in this paper.
Before proceeding, let us introduce some notation used in defining the
slip boundary conditions. We make use of a decomposition
of any vector $\vv$ on the boundary $\Gamma$ into its normal and tangential components,
$\vv = \vv_{\perp} + \vv_{\parallel}$, where $\vv_{\perp}=(\vv\cdot\nn)\nn$ and
$\vv_{\parallel}=P_{\nn}(\vv)$. The tangential projection is defined as
\begin{equation}
    P_{\nn}(\vv) = (\mathbf{I} - \nn\otimes\nn)\vv,
    \label{eq:projection_operator_appendix}
\end{equation}
where $\nn$ is the outer unit normal of the surface. Introducing
$\hat{\bsig} = \bsig\nn$ for the traction (vector), the forcing due
to cilia is represented by constraining $\bsigpar$ on the part of the
boundary populated by cilia. In particular, we let
$\bsigpar = \btau(\xx, \rr) = \tau\lambda(\xx) P_{\nn}(\rr)$,
where the parameter $\tau=\num{6.5e-4}$ Pa was used to control the magnitude
of the tangential stresses, and chosen by calibrating simulated velocity
fields with experimental data~\cite{Olstad2019CiliaryDevelopment}.
The details of the function $\lambda(\xx)$ follow subsequent to deriving the weak form.

We choose $\mathbf{r} = \rr^+ = (1, 0, 1)$ or $\mathbf{r} = \rr^- = -(1, 0, 1)$
to impose forces perpendicular to the $y$-axis.
The reason for this is that all of the experimental images we compare
with are taken in the $xz$-plane,
rendering flow features in the $y$-direction unobservable.
In the following, we consider the way of handling the stress
boundary integral for each model version successively.

@@

The boundary conditions~\Cref{eq:modelA_slip_BC} are set weakly through the stress
boundary integral 
\begin{equation}
    \intG{\bsig\nn\cdot\vv} =  \intG{\hat{\bsig}\cdot\vv}
    \label{eq:boundary_integral_BC1}
\end{equation}
in the weak form~\Cref{eq:modelA_weak_form1} of the Stokes equations.
We decompose the traction vector $\hat{\bsig}$ into a normal and tangential component:
\begin{equation*}
    \hat{\bsig} = \bsigperp + \bsigpar
\end{equation*}
Inserting the above into~\Cref{eq:boundary_integral_BC1},
combined with application of the tangential traction
condition~\Cref{eq:modelA_slip_BC_shear} and the
free-slip condition~\Cref{eq:modelA_full_slip_BC}, yields
\begin{equation}
    \intG{\hat{\bsig}\cdot\vv} = \int_{\Gc\cup\Gs}{(\bsigperp + \bsigpar)\cdot\vv}\, ds = 
    \int_{\Gc\cup\Gs}{\bsigperp\cdot\vv}\, ds
    + \intGc{\btau\cdot\vv} = \intG{\bsigperp\cdot\vv} + \intGc{\btau\cdot\vv}.
    \label{eq:boundary_integral_BC2}
\end{equation}
The second term on the right-hand side of~\Cref{eq:boundary_integral_BC2}
is moved to the right-hand side of the weak form, since $\btau$ is known
 Using $\hat{\bsig}_{\perp} = (\hat{\bsig}\cdot\nn)\nn$,
 we can write the first term on the right-hand side of~\Cref{eq:boundary_integral_BC2} as
\begin{equation}
    \intG{\bsigperp\cdot\vv} = \intG{((\hat{\bsig}\cdot\nn)\nn)\cdot\vv} =
    \intG{(\hat{\bsig}\cdot\nn)(\vv\cdot\nn)}.
    \label{eq:boundary_integral_BC_normal}
\end{equation}
Since $\Gp=\emptyset$ in this model version, and the term~\Cref{eq:boundary_integral_BC_normal}
vanishes owing to the construction $\VV$. As a result, for the cilia-only/no-cardiac model,
the weak form of the momentum equation takes the following form:
\begin{equation*}
    \intO{\rho\pdifft{\uu}\cdot\vv}
    + \intO{2\mu\beps(\uu) : \beps(\vv)}
    - \intO{p\nabla\cdot\vv} = \intGc{\btau\cdot\vv}, \quad\forall\vv\in\VV.
\end{equation*}
Next, multiplying the continuity equation~\Cref{eq:BDM_stokes_div}
with a test function $q\in Q$, we require that
\begin{equation}
    -\intO{q\nabla\cdot\uu} = 0,\quad\forall q\in Q
    \label{eq:pressure_weak_form}
\end{equation}
The full weak form of the Stokes equations for this flow model is then:
find $(\uu, \, p) \in \VV\times Q$, such that
\begin{subequations}
    \begin{alignat}{5}
        &\intO{\rho\pdifft{\uu}\cdot\vv}
        + &\intO{2\mu\beps(\uu) : \beps(\vv)}
        - &\intO{p\nabla\cdot\vv}
        &&= \intGc{\btau\cdot\vv}, &&\quad\forall\vv\in\VV, \label{eq:stokes_weak_mom_modelA}\\ 
        &{} &{}  -&\intO{q\nabla\cdot\uu} &&= 0, &&\quad\forall q\in Q.
        \label{eq:stokes_weak_cont_modelA}
    \end{alignat}%
    \label{eq:stokes_weak_modelA}%
\end{subequations}%
We observe that, in this case with slip boundary conditions on the entirety of
the boundary $\Gamma$, the vector $(\uu, p)=(\mathbf{0}, C)$ for any real
constant $C$ satisfies the equations.
The system~\Cref{eq:stokes_weak_modelA} is thus singular with a one-dimensional nullspace
of constant pressures.
Consequently, assembly of our discretization of the  equations~\Cref{eq:stokes_weak_modelA}
into a linear equation system $\mathbf{A}\mathbf{x}=\mathbf{b}$ yields a singular matrix $\mathbf{A}$.
To ensure existence of solutions, we orthogonalize the right-hand side vector $\mathbf{b}$ with
respect to the pressure nullspace basis (the set of all constants), before solving the linear
system. To obtain a unique solution, which together with existence renders the
problem well-posed, information of the pressure nullspace is passed to the linear solver.
Let us remark that the singularity is a consequence of the fact that the
impermeability condition is assumed to hold on the \emph{entire} domain boundary
for this model version, which leaves the pressure under-determined as a result of no
boundary condition specifying the pressure.

@@

\subsection{Discretization of the Stokes equations}
We will outline the discretization of the weak
formulation~\Cref{eq:abstract_weak_form_modelC,
eq:stokes_bilinear_form,eq:stokes_skew_form,eq:stokes_linear_form}
of the Stokes equations for the baseline model. A similar approach can
be used to discretize the weak formulation for the other model versions.
First, we consider the spatial discretization of the weak formulation.
We rewrite the weak form in terms of the discretized variables,
denoted with the subscript $h$, on the discrete domain $\Omega_h$:
\begin{subequations}
    \begin{alignat}{2}
        \mathcal{A}(\uu_h, \vv_h) + \mathcal{B}(\vv_h, p_h)
        &= \mathcal{L}(\vv_h), &&\quad\forall\vv_h\in\VV_h, \\
        \mathcal{B}(\uu_h, q_h) &= 0, &&\quad\forall q_h\in Q_h.
    \end{alignat}%
    \label{eq:no_stabilization_abstract_discrete_weak_form_modelC}%
\end{subequations}%
Our choice of discretization are finite elements
of the Brezzi-Douglas-Marini (BDM) family~\cite{Brezzi1985TwoProblems, Nedelec1986AIR3}
for the velocity
space and discontinuous Lagrange polynomial elements for the pressure. We refer to the latter
as discontinuous Galerkin (DG) elements. We define
\begin{align*}
    {\VV}_h &= \Big\{\vv_h : \vv_h\in\mathbf{H}(\mathrm{div});
    \vv_h\vert_K\in\big[P_k\big]^d;
    \forall K\in\Omega_h;
    \vv_h\cdot\nn=0 \ \text{on } \Gamma\setminus\Gp\Big\},\\
    Q_h &= \Big\{q_h : q_h\in L^2;  q_h\vert_K\in P_{k-1}; \forall K\in\Omega_h\Big\},
\end{align*}
where $P_k$ is the space of polynomials of degree less than or equal to $k$.
The BDM family of elements have degrees of freedom
associated with integral moments of vector components normal to facets of the mesh,
and is therefore non-conforming. However, the BDM-DG is $\mathbf{H}(\mathrm{div})$
conforming and thus exactly mass-conserving, because the continuity equation can be satisfied
in every computational mesh cell due to the fact that $\nabla\cdot\VV_h\subset Q_h$.
Moreover, control of normal components of $\uu_h$ conveniently 
lets us enforce the impermeability condition $\uu_h\cdot\nn=0$ strongly.
requires a stabilization of the bilinear form $\mathcal{A}(\uu_h, \vv_h)$.

Because the elements of $\VV_h$ define the normal components of $\uu_h$,
continuity in tangential components of $\uu_h$ on interior facets of
the mesh is ensured through stabilization
of the discrete weak form~\Cref{eq:no_stabilization_abstract_discrete_weak_form_modelC}.
We apply the stabilization outlined in Hong~\emph{et al.}~\cite{Hong2016AEquations}.
First, we define the average and jump operators, respectively, as
\begin{equation*}
    \avg{\uu} = \frac{\uu^+\cdot\nn^+ - \uu^-\cdot\nn^-}{2}
    \qquad\mathrm{and}\qquad \jump{\uu} = \uu^+ - \uu^-,
\end{equation*}
where $\nn$ is the normal vector on a facet separating two cells,
where the plus and minus signs denote variables in each of the two cells.
The normal vector is chosen such that it points towards the exterior of
its corresponding cell. On the boundary, the operators are defined as
$\avg{\uu} = \uu\cdot\nn$ and $\jump{\uu} = \uu$,
with $\nn$ pointing to the exterior of the domain. We now define
and add the bilinear form
\begin{align*}
    \mathcal{A}_{\mathrm{stab}}(\uu_h, \vv_h) &=
    \sum_{f\in \mathcal{F}}\int_f{2\mu\frac{\gamma}{\avg{h}}\jump{P_{\nn}(\uu_h)}
    \cdot\jump{P_{\nn}(\vv_h)}}\\
    &-\sum_{f\in \mathcal{F}}\int_f{2\mu\avg{\beps(\uu_h)}\cdot\jump{\vv_h}}
    -\sum_{f\in \mathcal{F}}\int_f{2\mu\avg{\beps(\vv_h)}\cdot\jump{\uu_h}}
\end{align*}
to the discrete weak formulation~\Cref{eq:no_stabilization_abstract_discrete_weak_form_modelC},
where $\mathcal{F}$ is the set of all interior facets of the computational mesh.
The parameter $\gamma>0$ penalizes jumps in the tangential components of the velocity
across facets, whereas $h$ is the cell diameter of a mesh cell,
defined as the maximal distance between two points in a cell.
The semi-discrete weak formulation of the Stokes equations reads:
find ($\uu_h, p_h$)$\in$($\VV_h, Q_h$) such that
\begin{subequations}
    \begin{alignat}{2}
        \mathcal{A}(\uu_h, \vv_h)
        + \mathcal{A}_{\mathrm{stab}}(\uu_h, \vv_h)
        + \mathcal{B}(\vv_h, p_h)
        &= \mathcal{L}(\vv_h), &&\quad\forall\vv_h\in\VV_h, \\
        \mathcal{B}(\uu_h, q_h) &= 0, &&\quad\forall q_h\in Q_h.
    \end{alignat}%
    \label{eq:abstract_semi_discrete_weak_form_modelC}%
\end{subequations}%
We solved~\Cref{eq:abstract_semi_discrete_weak_form_modelC}
numerically with the penalty parameter $\gamma=10$.

>>>>

\subsection{The role of motile cilia in intraventricular flow and transport}
Our work has studied CSF flow and solute distribution
in a relatively small and simple brain model, which is devoid of
active CSF secretion and CSF absorption. Even though there is
a clear conservation of the ventricular system across vertebrates,
there are also important considerations to take. These include (but are not limitied to)
ventricular size and morphology, properties of ciliated cells,
animal posture, and magnitude of velocities. For instance, a large
volume would require an extensive amount of time for solutes to distribute
if based solely on diffusion. Besides, the presence of multiple cilia
instead of a single cilium per cell -- the former can be found in
mammalian ependymal cells -- will generate stronger velocities.
Hence advection may be more dominant in those situations.

In large brains such as the human brain, we presume ciliary beating mostly 
contributes to flow in localized regions proximal to the ventricular walls due to
the microscopic length of the cilia.
Although the precise mechanisms of intraventricular flow are still debated,
much evidence indicate that blood vessel pulsations, respiration, CSF secretion
and brain motion induce the pulsatile bulk flow in the central
regions of the ventricles~\cite{Vinje2019RespiratoryMeasurements,
Siyahhan2014FlowVentricles, Linninger2005PulsatileBrain, Enzmann1992BrainImaging,
Kurtcuoglu2007ComputationalSylvius}.
In human brains, cilia therefore probably play an important role in regulating the
amount of substances like exosomes close to the ventricular walls, similar to coral
reefs, where it has been observed that cilia are primary regulators of
oxygen amounts at the coral surface~\cite{Pacherres2022CiliaryProduction}.
Flow mediated by cilia indeed serves such a purpose in nodal flow at early
developmental stages of the brain~\cite{Hirokawa2006NodalAsymmetry}, and the discovery
of a cilia-driven flow network in rodent brains suggest such selective
transport~\cite{Faubel2016Cilia-basedVentricles}.

On the other hand, for the zebrafish embryos we studied,
the ventricles are so small that the flow
generated by ciliary motion is of the scale of the geometry, and thus not
only contributes to near-wall flow dynamics, but flow in the entire ventricular
system~\cite{Olstad2019CiliaryDevelopment}.
To date, motile-cilia deficiency has been associated with altered brain function
in the zebrafish, dysfunctional establishment of left-right symmetry,
and ventricular enlargement in zebrafish, rodents and humans.~\cite{Afzelius2004Cilia-relatedDiseases,
Youn2018PrimaryDiseases, Ma2024CiliaDisease, Ibanez-Tallon2002LossHydrocephalus}
Yet the molecular mechanisms leading to these alterations are poorly understood.
We trust that our model sets the basis for future work addressing these open questions.
We are also confident that our model can be adjusted and implemented to fit
microscopic and macroscopic parameters for larger brains, once these parameters are acquired.

\subsection{Limitations and further work}
The model presented has certain limitations. Future work could consider:
\begin{itemize}
    \item spatial heterogeneity of the cilia, how this affects flow and transport
    \item time-varying cilia?
    \item improved advection-diffusion equation influx/outflux BCs, or instead
    modeling deformation of a closed ventricular system. 
    and investigating how the arising flow compares with pulsatile flow observed in experiments
    \item FSI: CSF flow coupled to 1D/2D/3D cilia structures.
\end{itemize}

\section{Conclusion}
We have presented a numerical model
of cerebrospinal fluid flow and solute transport in
zebrafish brain ventricles based on finite element methods.
A realistic ventricular geometry from medical imaging was used to perform a
computational study. The flow model reproduces
the experimentally observed flow features of cilia-mediated cerebrospinal fluid
flow in zebrafish brain ventricles. Coupling the flow to the transport model,
we estimated how slowly or rapidly molecules of varying size are transported in
the ventricles. 
The numerical model was validated with in vivo experiments, measuring the movement of
a photoconverted fluorescent protein in zebrafish brain ventricles,
both for healthy animals and mutant animals with ciliary dysfunction.
We observed good alignment between computational and experimental results.
Simulations indicate that cilia-driven advection is important
for transport of larger molecules, whereas diffusion accounts for most of
the transport of smaller molecules. 
Experimental results support the conclusion that cilia only affect the
local distribution of small proteins,
while most of their transport is governed by diffusion.

@@

Further steps depend on the boundary conditions applied,
which is different for the three model versions we consider in this paper.
Before proceeding, let us introduce some notation used in defining the
slip boundary conditions. We make use of a decomposition
of any vector $\vv$ on the boundary $\Gamma$ into its normal and tangential components,
$\vv = \vv_{\perp} + \vv_{\parallel}$, where $\vv_{\perp}=(\vv\cdot\nn)\nn$ and
$\vv_{\parallel}=P_{\nn}(\vv)$. The tangential projection is defined as
\begin{equation}
    P_{\nn}(\vv) = (\mathbf{I} - \nn\otimes\nn)\vv,
    \label{eq:projection_operator_appendix}
\end{equation}
where $\nn$ is the outer unit normal of the surface. Introducing
$\hat{\bsig} = \bsig\nn$ for the traction (vector), the forcing due
to cilia is represented by constraining $\bsigpar$ on the part of the
boundary populated by cilia. In particular, we let
$\bsigpar = \btau(\xx, \rr) = \tau\lambda(\xx) P_{\nn}(\rr)$,
where the parameter $\tau=0.65$ mPa was used to control the magnitude
of the tangential stresses, and chosen by calibrating simulated velocity
fields with experimental data~\cite{Olstad2019CiliaryDevelopment}.
The details of the function $\lambda(\xx)$ follow subsequent to deriving the weak form.

We choose $\rr = \rr^+ = (1, 0, 1)$ or $\rr = \rr^- = -(1, 0, 1)$
to impose forces perpendicular to the $y$-axis.
The reason for this is that all of the experimental images we compare
with are taken in the $xz$-plane,
rendering flow features in the $y$-direction unobservable.
In the following, we consider the way of handling the stress
boundary integral for each model version successively.

@@

The boundary conditions~\Cref{eq:modelA_slip_BC} are set weakly through the stress
boundary integral 
\begin{equation}
    \intG{\bsig\nn\cdot\vv} =  \intG{\hat{\bsig}\cdot\vv}
    \label{eq:boundary_integral_BC1}
\end{equation}
in the weak form~\Cref{eq:modelA_weak_form1} of the Stokes equations.
We decompose the traction vector $\hat{\bsig}$ into a normal and tangential component:
\begin{equation*}
    \hat{\bsig} = \bsigperp + \bsigpar
\end{equation*}
Inserting the above into~\Cref{eq:boundary_integral_BC1},
combined with application of the tangential traction
condition~\Cref{eq:modelA_slip_BC_shear} and the
free-slip condition~\Cref{eq:modelA_full_slip_BC}, yields
\begin{equation}
    \intG{\hat{\bsig}\cdot\vv} = \int_{\Gc\cup\Gs}{(\bsigperp + \bsigpar)\cdot\vv}\, ds = 
    \int_{\Gc\cup\Gs}{\bsigperp\cdot\vv}\, ds
    + \intGc{\btau\cdot\vv} = \intG{\bsigperp\cdot\vv} + \intGc{\btau\cdot\vv}.
    \label{eq:boundary_integral_BC2}
\end{equation}
The second term on the right-hand side of~\Cref{eq:boundary_integral_BC2}
is moved to the right-hand side of the weak form, since $\btau$ is known
 Using $\hat{\bsig}_{\perp} = (\hat{\bsig}\cdot\nn)\nn$,
 we can write the first term on the right-hand side of~\Cref{eq:boundary_integral_BC2} as
\begin{equation}
    \intG{\bsigperp\cdot\vv} = \intG{((\hat{\bsig}\cdot\nn)\nn)\cdot\vv} =
    \intG{(\hat{\bsig}\cdot\nn)(\vv\cdot\nn)}.
    \label{eq:boundary_integral_BC_normal}
\end{equation}
Since $\Gp=\emptyset$ in this model version, and the term~\Cref{eq:boundary_integral_BC_normal}
vanishes owing to the construction $\VV$. As a result, for the cilia-only/no-cardiac model,
the weak form of the momentum equation takes the following form:
\begin{equation*}
    \intO{\rho\pdifft{\uu}\cdot\vv}
    + \intO{2\mu\beps(\uu) : \beps(\vv)}
    - \intO{p\nabla\cdot\vv} = \intGc{\btau\cdot\vv}, \quad\forall\vv\in\VV.
\end{equation*}
Next, multiplying the continuity equation~\Cref{eq:BDM_stokes_div}
with a test function $q\in Q$, we require that
\begin{equation}
    -\intO{q\nabla\cdot\uu} = 0,\quad\forall q\in Q
    \label{eq:pressure_weak_form}
\end{equation}
The full weak form of the Stokes equations for this flow model is then:
find $(\uu, \, p) \in \VV\times Q$, such that
\begin{subequations}
    \begin{alignat}{5}
        &\intO{\rho\pdifft{\uu}\cdot\vv}
        + &\intO{2\mu\beps(\uu) : \beps(\vv)}
        - &\intO{p\nabla\cdot\vv}
        &&= \intGc{\btau\cdot\vv}, &&\quad\forall\vv\in\VV, \label{eq:stokes_weak_mom_modelA}\\ 
        &{} &{}  -&\intO{q\nabla\cdot\uu} &&= 0, &&\quad\forall q\in Q.
        \label{eq:stokes_weak_cont_modelA}
    \end{alignat}%
    \label{eq:stokes_weak_modelA}%
\end{subequations}%
We observe that, in this case with slip boundary conditions on the entirety of
the boundary $\Gamma$, for any real constant $C$ the vector
$(\uu, p)=(\mathbf{0}, C)$ satisfies the equations.
Hence the system~\Cref{eq:stokes_weak_modelA} is singular with a one-dimensional nullspace
of constant pressures.
Consequently, assembly of our discretization of the  equations~\Cref{eq:stokes_weak_modelA}
into a linear equation system $\mathbf{A}\mathbf{x}=\mathbf{b}$ yields a singular matrix $\mathbf{A}$.
To ensure existence of solutions, we orthogonalize the right-hand side vector $\mathbf{b}$ with
respect to the pressure nullspace basis (the set of all constants), before solving the linear
system. To obtain a unique solution, which together with existence renders the
problem well-posed, information of the pressure nullspace is passed to the linear solver.
Let us remark that the singularity is a consequence of the fact that the
impermeability condition is assumed to hold on the \emph{entire} domain boundary
for this model version, which leaves the pressure under-determined as a result of no
boundary condition specifying the pressure.

@@

\subsection{Discretization of the Stokes equations}
We will outline the discretization of the weak
formulation~\Cref{eq:abstract_weak_form_modelC,
eq:stokes_bilinear_form,eq:stokes_skew_form,eq:stokes_linear_form}
of the Stokes equations for the baseline model. A similar approach can
be used to discretize the weak formulation for the other model versions.
First, we consider the spatial discretization of the weak formulation.
We rewrite the weak form in terms of the discretized variables,
denoted with the subscript $h$, on the discrete domain $\Omega_h$:
\begin{subequations}
    \begin{alignat}{2}
        \mathcal{A}(\uu_h, \vv_h) + \mathcal{B}(\vv_h, p_h)
        &= \mathcal{L}(\vv_h), &&\quad\forall\vv_h\in\VV_h, \\
        \mathcal{B}(\uu_h, q_h) &= 0, &&\quad\forall q_h\in Q_h.
    \end{alignat}%
    \label{eq:no_stabilization_abstract_discrete_weak_form_modelC}%
\end{subequations}%
Our choice of discretization are finite elements
of the Brezzi-Douglas-Marini (BDM) family~\cite{Brezzi1985TwoProblems, Nedelec1986AIR3}
for the velocity
space and discontinuous Lagrange polynomial elements for the pressure. We refer to the latter
as discontinuous Galerkin (DG) elements. For all cells $K$ of the computational mesh $\Omega_h$,
we define
\begin{align*}
    {\VV}_h &= \Big\{\vv_h : \vv_h\in\mathbf{H}(\mathrm{div});
    \vv_h\vert_K\in\big[P_k\big]^d;
    \forall K\in\Omega_h;
    \vv_h\cdot\nn=0 \ \text{on } \Gamma\setminus\Gp\Big\},\\
    Q_h &= \Big\{q_h : q_h\in L^2;  q_h\vert_K\in P_{k-1}; \forall K\in\Omega_h\Big\},
\end{align*}
where $P_k$ is the space of polynomials of degree less than or equal to $k$.
The BDM family of elements have degrees of freedom
associated with integral moments of vector components normal to facets of the mesh,
and is therefore non-conforming. However, the BDM-DG is $\mathbf{H}(\mathrm{div})$
conforming and thus exactly mass-conserving, because the continuity equation can be satisfied
in every computational mesh cell due to the fact that $\nabla\cdot\VV_h\subset Q_h$.
Moreover, control of normal components of $\uu_h$ conveniently 
lets us enforce the impermeability condition $\uu_h\cdot\nn=0$ strongly.
requires a stabilization of the bilinear form $\mathcal{A}(\uu_h, \vv_h)$.

Because the elements of $\VV_h$ define the normal components of $\uu_h$,
continuity in tangential components of $\uu_h$ on interior facets of
the mesh is ensured through stabilization
of the discrete weak form~\Cref{eq:no_stabilization_abstract_discrete_weak_form_modelC}.
We apply the stabilization outlined in Hong~\emph{et al.}~\cite{Hong2016AEquations}.
First, we define the average and jump operators, respectively, as
\begin{equation*}
    \avg{\uu} = \frac{\uu^+\cdot\nn^+ - \uu^-\cdot\nn^-}{2}
    \qquad\mathrm{and}\qquad \jump{\uu} = \uu^+ - \uu^-,
\end{equation*}
where $\nn$ is the normal vector on a facet separating two cells,
where the plus and minus signs denote variables in each of the two cells.
The normal vector is chosen such that it points towards the exterior of
its corresponding cell. On the boundary, the operators are defined as
$\avg{\uu} = \uu\cdot\nn$ and $\jump{\uu} = \uu$,
with $\nn$ pointing to the exterior of the domain. We now define
and add the bilinear form
\begin{align*}
    \mathcal{A}_{\mathrm{stab}}(\uu_h, \vv_h) &=
    \sum_{f\in \mathcal{F}}\int_f{2\mu\frac{\gamma}{\avg{h}}\jump{P_{\nn}(\uu_h)}
    \cdot\jump{P_{\nn}(\vv_h)}}\\
    &-\sum_{f\in \mathcal{F}}\int_f{2\mu\avg{\beps(\uu_h)}\cdot\jump{\vv_h}}
    -\sum_{f\in \mathcal{F}}\int_f{2\mu\avg{\beps(\vv_h)}\cdot\jump{\uu_h}}
\end{align*}
to the discrete weak formulation~\Cref{eq:no_stabilization_abstract_discrete_weak_form_modelC},
where $\mathcal{F}$ is the set of all interior facets of the computational mesh.
The first term of $\mathcal{A}_{\mathrm{stab}}$ penalizes jumps
in the tangential components of the velocity
across interior facets, where the penalty parameter $\gamma>0$ must be chosen large enough
to achieve stability. For convergence the penalization term
scales with the cell diameter of a mesh cell $h$,
defined as the maximal distance between two points in a cell. The other two terms
of $\mathcal{A}_{\mathrm{stab}}$ are added to ensure coercivity and symmetry of
the bilinear form, such that the problem is well-defined.

Including the stabilization outlined above, the semi-discrete weak formulation of
the Stokes equations reads:
find ($\uu_h, p_h$)$\in$($\VV_h, Q_h$) such that
\begin{subequations}
    \begin{alignat}{2}
        \mathcal{A}(\uu_h, \vv_h)
        + \mathcal{A}_{\mathrm{stab}}(\uu_h, \vv_h)
        + \mathcal{B}(\vv_h, p_h)
        &= \mathcal{L}(\vv_h), &&\quad\forall\vv_h\in\VV_h, \\
        \mathcal{B}(\uu_h, q_h) &= 0, &&\quad\forall q_h\in Q_h.
    \end{alignat}%
    \label{eq:abstract_semi_discrete_weak_form_modelC}%
\end{subequations}%
We solved~\Cref{eq:abstract_semi_discrete_weak_form_modelC}
numerically with the penalty parameter $\gamma=10$.
